A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility

### 2019-06-17 paper.Rmd
misc changes to section 2 on challanges/tradeoffs

Data are a fundamental currency upon which scientific discoveries are made. Without access to good data, it becomes extremely difficult, if not impossible, to advance science. Yet, a large majority of data on which published research papers are based rarely see the light of day and are only visible to the original authors [@Rowhani-Farid2016; @Stodden2018].  Sharing data sets upon publication of a research paper has benefits to individual researchers, often through increased visibility of the work [@ Popkin2019; @Kirk2019]. But there is also a component of benefit to the broader scientific community. This primarily comes in the form of potential for reuse in other contexts along use for training and teaching (@McKiernan2016). Assuming that the data have no privacy concerns (e.g., human subjects, locations of critically endangered species), or that the act of sharing doesn't put the authors at a competitive disadvantage, sharing data will always have a net positive benefit. First and foremost, sharing data along with other artifacts can make it easier for a reader to independently reproduce the results, thereby increasing transparency and trust in the work. The act of easy data sharing can also improves model training, as many different models can be tried and tested on latest data sources, closing the loop on research and application of statistical techniques. Existing data sets can be combined or linked with new or existing data, fostering the development and synthesis of new ideas and research areas. The biggest of these benefits is the overall increase in reproducibility.

@@

A typical data scientist loads a dozen or two of these open source libraries at the top of a notebook and then rely on existing routines to rapidly read, transform, visualize, and model data. Each of these packages individually depend on a complex web of other packages, building upon existing work rather than re-implementing everything from scratch. Working from script and a list of such dependencies, a data analyst can easily install all the necessary tools in any local or remote environment and reproduce the computation. When new functionality is developed, it is packaged into a separate entity and added to a language's package manager.

@@

Although there are numerous public and private data repositories, none  of them function as package managers. Datasets are also far more diverse in complexity, size, and formats, making them particularly challenging to standardize or easily "install" where the code is running. As a result, data used in an analysis is often read from various locations (local, network, or cloud), various formats, varying levels of tidiness [@Wickham2014]. There is also the overhead associated with data publishing, the act of archiving data in a repository that also mints permanent identifiers, that are not required of all projects due to the effort and time involved. It is worth drawing the distinction between data sharing (making the data available with little effort) and data publishing (archiving the data in a long-term repository, with our without curated metadata).

@@

@@

Providing a well documented dataset alongside research requires meeting several challenges. This includes understanding all the pieces and features of documentation such as metadata, licenses, and raw data; knowing where to share the data to ensure long term storage.

Better documentation requires more time. Higher impact data needs better documentation.

It takes more effort to prepare data to make it easy to understand and share. The challenge here is understanding how much time to spend to make it easier to understand before the gains in understanding are outweighed by the time taken to improve them.

There is also a challenge in understanding the impact of data. Data with high potential for impact should be made easier to understand and share. This can be seen in the research created from the 50 hectare plot, discussed in section ... - this dataset was challenging to understand without special domain expertise.

Simpler format data like CSV is more reliable than binary serialised/compressed data (ask Laura DeCicco re this).

Providing the right amount of documentation for your data is challenging. Too much is overwhelming; too little and you don't know anything. For example, compare a folder with dozens of data files and several papers, to a single CSV file. The first is too much, and understanding it requires a high time cost. The second is too little; we don't know anything about what the data is or why it was collected. The challenge is to provide the right level of detail, somewhere between these two extremes.

There are many features to consider when documenting data, such as a README file with human-readable descriptions of data and variables, data licenses, whether to include raw data, data cleaning scripts, and machine readable metadata. Together these all help make the data easier to understand. These features are not clearly described in the scientific literature in the context of including them alongside research.

To frame discussion around the challenges of data documenting, we can think of two features: "Effort to prepare", and "Ease of understanding". We can plot these on two axes and place datasets along each to indicate the ease of understanding and preparation (see Figure \@ref(fig:fig-effort-understanding)).

@@

## Higher potential impact means more time

To improve data impact it needs more time to prepare so it is easier to understand. Accordingly, data with higher potential for impact and reuse should be made as easy to understand as possible. Therefore data with high potential impact requires more time and effort to prepare. For example, data from medical randomized control trials (RCTs) on cancer treatment can have high impact, so requires a lot of time and effort document. Comparatively, a small survey on simple demographics can have low impact. So it might not have a worthwhile tradeoff for ease of understanding and ease of preparation. This means effort put into preparing data documentation for a small survey to be made easy to use should be kept relatively simple and not over complicated.

@@

## Understanding tradeoffs

@@

# Conclusions

@@

>>>>

Data are a fundamental currency upon which scientific discoveries are made. Without access to good data, it becomes extremely difficult, if not impossible, to advance science. Yet, a large majority of data on which published research papers are based rarely see the light of day and are only visible to the original authors [@Rowhani-Farid2016; @Stodden2018].  Sharing data sets upon publication of a research paper has benefits to individual researchers, often through increased visibility of the work [@Popkin2019; @Kirk2019]. But there is also a component of benefit to the broader scientific community. This primarily comes in the form of potential for reuse in other contexts along use for training and teaching (@McKiernan2016). Assuming that the data have no privacy concerns (e.g., human subjects, locations of critically endangered species), or that the act of sharing doesn't put the authors at a competitive disadvantage, sharing data will always have a net positive benefit. First and foremost, sharing data along with other artifacts can make it easier for a reader to independently reproduce the results, thereby increasing transparency and trust in the work. The act of easy data sharing can also improves model training, as many different models can be tried and tested on latest data sources, closing the loop on research and application of statistical techniques. Existing data sets can be combined or linked with new or existing data, fostering the development and synthesis of new ideas and research areas. The biggest of these benefits is the overall increase in reproducibility.

@@

A typical data analysis loads a dozen or two of these open source libraries at the top of a notebook and then relies on existing routines to rapidly read, transform, visualize, and model data. Each package depends on a complex web of other packages, building upon existing work rather than re-implementing everything from scratch. Working from script and a list of such dependencies, a data analyst can easily install all the necessary tools in any local or remote environment and reproduce the computation. When new functionality is developed, it is packaged into a separate entity and added to a language's package manager.

@@

Datasets are often far more diverse than code in terms of complexity, size, and formats. This makes them particularly challenging to standardize or easily "install" where the code is running. While there are numerous public and private data repositories, none of them function as package managers, which is what provides so much robustness to code sharing. As a result, data used in an analysis is often read from various locations (local, network, or cloud), various formats, varying levels of tidiness [@Wickham2014]. There is also the overhead associated with data publishing, the act of archiving data in a repository that also mints permanent identifiers, that are not required of all projects due to the effort and time involved. It is worth drawing the distinction between data sharing (making the data available with little effort) and data publishing (archiving the data in a long-term repository, with our without curated metadata).

@@

There are many features to consider when documenting, sharing, and publishing data. What format to store the data in? How do you describe the variables, observations, and collection methods? Do you include raw data? What about data used in analysis? What about metadata for the variables? And then there's linceses, and long term storage. It can be hard to know where to start. This section discusses challenges that come with these features of data documenting, sharing, and publication, setting up the discussion for the rest of the paper.

## How far to go with data documenting

To frame discussion around the challenges of data documenting, we can think of how an individual dataset fits on two features: "Effort to prepare", and "Ease of understanding", see Figure \@ref(fig:fig-effort-understanding).

@@

@@

There are many features to include in data documentation.
Not all of these are needed for every dataset. But, deciding which to include should be based upon the potential impact of the data. The more impactful the data, the more features to include. This is one of the first challanges (tradeoffs) to discuss: how to decide what to include with your data.

Together these all help make the data easier to understand. These features are not clearly described in the scientific literature in the context of including them alongside research.

## Data format

There is also the challenge for format. A simpler format data like CSV (Comma Separated Value) is more reliable than binary serialised or compressed data, but can sometimes be slower to read in and out, and harder to share.

To improve data impact it needs more time to prepare so it is easier to understand. Accordingly, data with higher potential for impact and reuse should be made as easy to understand as possible. Therefore data with high potential impact requires more time and effort to prepare. For example, data from medical randomized control trials (RCTs) on cancer treatment can have high impact, so requires a lot of time and effort document. Comparatively, a small survey on simple demographics can have low impact. So it might not have a worthwhile tradeoff for ease of understanding and ease of preparation. This means effort put into preparing data documentation for a small survey to be made easy to use should be kept relatively simple and not over complicated. Likewise, data that can be created via simulation from open source software could arguably not be archived - since it can be generated from scratch with code; a reproducible process with likely very low cost in person time to create.

@@

High impact data should be easy to understand and use, but how do you assess its impact?

## Describing the variables and metadata

## Including raw data

## Data used in analysis

## Licenses

## Long term storage

## Just how much to include

Another challenge is in the gross total amount of documentation. Too much documentation can be overwhelming; too little and you don't know anything. For example, compare a folder with dozens of data files and several papers, to a single CSV file. The first is too much, and understanding it requires a high time cost. The second is too little; we don't know anything about what the data is or why it was collected. The challenge is to provide the right level of detail, somewhere between these two extremes. This changes depending on the impact of the data, and the task at hand.

Creating good documentation has the same challenge as doing good writing: it takes time, and it can be hard to know when you are done.

@@

@@

# Morgue

## Challenges -  Understanding tradeoffs

The research created from the 50 hectare plot, discussed in section \@ref(ex-data-1) - this dataset was challenging to understand without special domain expertise.

Ultimately, creating good documentation has the same challenge as good writing: it takes time, and it can be hard to know when you are done.

It can be hard to find the right amount of time to spend. Ideally, you want the gains in understanding not to be outweighed by the time taken to improve them.

Impact is a difficult quantity to measure.
